Classification and characterization of anthropogenic plant communities in the northwestern Iberian Peninsula

### 2024-10-04 doc/manuscript.Rmd
Revised classification, no Aegopodion

To assist in our classification, we prepared a checklist of anthropogenic syntaxa that could be present in our study area (**Appendix S1**), based on regional syntaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. This syntaxonomical checklist included 39 anthropogenic vegetation alliances that could be present in the Iberian Atlantic territories according to the literature.

@@

We started this stage of the analyses by keeping only those plots (n = 2,218) that had been assigned by the original authors of the plot to any of the 39 anthropogenic alliances defined in our syntaxonomical checklist (i.e., we removed the plots without syntaxa, that had been retrieved solely by the expert system). We further cleaned this subset using TWINSPAN classification (with the same settings as above), in which we attempted to match the plots into the 39 alliances. In a first round of TWINSPAN, we divided the dataset into an increasing number of clusters, until reaching 15 clusters, beyond which the new created clusters were not phytosociologically meaningful. Then, we performed a second round of TWINSPAN within each of the 15 clusters. In this second round, we assigned each of the clusters to an alliance in our checklist, based on the composition of diagnostic species as described in the regional literature. Within each of the 15 primary clusters, we created an increasing number of secondary clusters until the newly created clusters splited one of the alliances that had already been assigned.

This stage of the analyses allowed us to identify those plots in which there was an agreement between (1) our TWINSPAN-based classification (i.e. numerical classification) and (2) the classification based on the original author's syntaxon (i.e. regional phytosociological expertise). These plots (n = 1,725) became our training dataset for semi-supervised classification. Furthermore, this step allowed us to validate which alliances from our syntaxonomical checklist were present in the dataset (n = 28) and which were absent (n = 10).

@@

Next, we conducted a semi-supervised classification [@RN5593] of the whole dataset (n = 2,508 plots) into the 28 validated anthropogenic alliances. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Since our goal was to refine the classification of the whole dataset, we allowed plots from the training subset to be re-assigned to other alliances during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups, but attempts to do so resulted in new groups with no ecological significance, and thus we kept the 28 alliances as final anthropogenic vegetation groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows plots to be classified either into the *a priori* alliances or into a *noise* group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package *vegclust* [@RN5383]. The semi-supervised classification with noise clustering resulted in the final classification of 2,081 plots into 28 alliances and 427 plots left out in the noise group. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances, and Holm's p-value correction) fitted with the R package *RVAideMemoire* [@RN5381] to test the significance of the final vegetation alliances; along with Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] to visualize the relationships between the classes and alliances.

>>>>

To assist in our classification, we prepared a checklist of anthropogenic syntaxa that could be present in our study area (**Appendix S1**), based on regional syntaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. This syntaxonomical checklist included 36 anthropogenic vegetation alliances that could be present in the Iberian Atlantic territories according to the literature.

@@

We started this stage of the analyses by keeping only those plots (n = 2,100) that had been assigned by the original authors of the plot to any of the 39 anthropogenic alliances defined in our syntaxonomical checklist (i.e., we removed the plots without syntaxa, that had been retrieved solely by the expert system). We further cleaned this subset using TWINSPAN classification (with the same settings as above), in which we attempted to match the plots into the 39 alliances. In a first round of TWINSPAN, we divided the dataset into an increasing number of clusters, until reaching 15 clusters, beyond which the new created clusters were not phytosociologically meaningful. Then, we performed a second round of TWINSPAN within each of the 15 clusters. In this second round, we assigned each of the clusters to an alliance in our checklist, based on the composition of diagnostic species as described in the regional literature. Within each of the 15 primary clusters, we created an increasing number of secondary clusters until the newly created clusters splited one of the alliances that had already been assigned.

This stage of the analyses allowed us to identify those plots in which there was an agreement between (1) our TWINSPAN-based classification (i.e. numerical classification) and (2) the classification based on the original author's syntaxon (i.e. regional phytosociological expertise). These plots (n = 1,767) became our training dataset for semi-supervised classification. Furthermore, this step allowed us to validate which alliances from our syntaxonomical checklist were present in the dataset (n = 27) and which were absent (n = 9).

## Not supported by TWINSPAN

*Caucalidion lappulae* (4)
*Roemerion hybridae* (2)
*Hordeion murini* (2)
*Taeniathero-Aegilopion geniculatae* (1)
*Onopordion castellani* (6)
*Euphorbion prostratae* (4)

*Onopordion acanthii* (5)
*Fragarion vescae* (1)
*Aegopodion podagrariae* (39)

@@

Next, we conducted a semi-supervised classification [@RN5593] of the whole dataset (n = 2,508 plots) into the 28 validated anthropogenic alliances. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Since our goal was to refine the classification of the whole dataset, we allowed plots from the training subset to be re-assigned to other alliances during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups, but attempts to do so resulted in new groups with no ecological significance, and thus we kept the 28 alliances as final anthropogenic vegetation groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows plots to be classified either into the *a priori* alliances or into a *noise* group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package *vegclust* [@RN5383]. The semi-supervised classification with noise clustering resulted in the final classification of 2,050 plots into 27 alliances and 458 plots left out in the noise group. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances, and Holm's p-value correction) fitted with the R package *RVAideMemoire* [@RN5381] to test the significance of the final vegetation alliances; along with Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] to visualize the relationships between the classes and alliances.
